

# The "Independent" newspaper tabulated the gender of all candidates running for the 1992 British general election (data set "belection.rda"). 
# Find a suitable glm, ensuring a diagnostic analysis, and intepretation of the model.

Data contains 2 instances of northwest, the second one of each should be simply north 
```{r}
load("belection.rd")

levels(belection$Region) <- c(levels(belection$Region), 'North')

belection$Region[c(11,22,33,44,55)] <- "North"

belection

```


Appears to be a compressed data frame binomial problem with y = 1 being females (the interesting value) with two explanatory variables party and region.

11 regions {EastAnglia, EastMidlands, GreaterLondon, NorthWest, North, Scotland, SouthEast, SouthWest, Wales, WestMidlands, YorkHumbers}

5 parties {Cons, Green, Labour, LibDem, Other}

For which, each combination provides a count of males and females candidates for the 1992 British general election.



# Exploratory Data Analysis
proportion of females against party

```{r}
plot(Females/(Females+Males) ~ Party, data=belection)

```
# Comments 
The proportion of females going for a electoral positions is lowest for the cons, which have the narrowest distribution shown from the bar plots above. 
The green party has the highest mean at around a proportion of 0.27 and the widest distribution, but also have an outlier with the smallest values (0 proportion of women)
The other three parties have similar mean values around 0.2, with the highest distribution being in other, then labour party having the widest distribution, and libdem having a tighter range. 
The differing parties will play a role in determining the proportion of females running for election, but the highest influence will be by Cons as the other groups are quite similar. 


# Female candidates against region 

```{r}
plot(Females/(Females+Males) ~ Region, data=belection, xaxt= "n", xlab = "")
axis(side = 1, at= 1:11, las = 2, label = levels(belection$Region))

```
# Comment
There is a differing dispersion between the differing areas, with an overall mean around 0.15-0.25.
The lowest mean is in Wales and includes the only 0 result.
The highest mean is at SouthEast, closely followed by south west. South east is also quite consistent as it has a small box, with only one downward outlier. South west dispersion is much higher. 
The highest value can be seen as an outlier for East Anglia at 0.4
The second highest point can be seen in the North, which also has the lowest upper quartile range and second lowest mean, with two outliers one higher one lower. 
From this the location appears to be important in determining the proportion of females that ran for election. 
Assumption of a binomial GLM : individuals in each group are assumed to be independent.

# Model

Binomial glm 

log(odds) = B0 + B1x1 

We want to discover if Region and Party have an impact on the proportion of female candidates.

We also looked at interaction Region * Party, but Region was not significant from the Anova test.

```{r}

be.m1 <-  glm(cbind(Females,Males) ~  Party + Region, family=binomial, data=belection)

summary(be.m1)

anova(be.m1, test = "Chisq")

```

# Comment 
- Region is not significant 


We want to discover if Party have an impact on the proportion of female candidates.

```{r}

be.m2 <-  glm(cbind(Females,Males) ~  Party , family=binomial, data=belection)

summary(be.m2)

anova(be.m2, test = "Chisq")

exp( coef(be.m2) )


library(statmod)
qres <- qresid(be.m2); qqnorm(qres, las=1); abline(0, 1)
scatter.smooth( qres~fitted(be.m2), las=1, main="Residuals vs fitted",
xlab="Fitted value", ylab="Quantile residual")

```
The model appears to be good, unexplained deviance appears to be a bit high.

Plot of residual vs fitted values (well-behaved plot):
-residuals are randomly around the 0 line so the assumption of linear relationship is reasonable
-residuals roughly form an horizontal band around the 0 line, this suggests that the variances of the error term are equal
-no residual stands out from the basic random pattern of residuals, this suggests that there are no outliers


# Broken assumptions:
From inspecting the data it can be seen that the saddle point and central limit theorem assumptions may be violated for small dispersion asymptopics:

the saddle point for a binomial is my >= 3 and m(1-y) >= 3 
and for central limit theory my >= 5 and m(1-y) >= 5

Unfortunately there are point that violate this within the data set including a 1, 0 and 2, as well as a host of threes. 

Can look to combine some of the regions to increase the counts, north was one of the main causes.

First of all the data set already has already been altered to have 2 northwest variables instead of one north and one not, by combining these point will result in a dataframe with 50 obs.


# Alternatively can combine north and northwest - increasing the minimum count 

```{r}
library(dplyr)
load("belection.rd")


belection_2 <- belection %>%
  group_by(Party, Region) %>%
  summarise(across(c(Males, Females), sum))

belection_2
```

We consider that some regions have only few observations, like East Anglia has 1 observation and Wales has 2 and a 0.

This link shows a map of Uk regions:
https://www.alamy.com/east-midlands-red-highlighted-in-map-of-england-uk-image274693275.html?imageid=2C1A37B2-84C1-4C1F-B48B-B3554E76ECD2&p=347647&pn=1&searchId=4d9274a70f0c04a6c9026ec96c704868&searchtype=0

We have already joined Northwest and north east according to the map.
Joins of interest are west midlands with Wales, and East Anglia with East Midlands.


```{r}
for (i in 1:length(belection_2$Region)){
  
  if(belection_2$Region[i] == "Wales") 
  {belection_2$Region[i] = "WestMidlands"}
  
  if(belection_2$Region[i] == "EastAnglia") 
  {belection_2$Region[i] = "EastMidlands"}
  
}

belection_2 <- belection_2 %>%
  group_by(Party, Region) %>%
  summarise(across(c(Males, Females), sum))

belection_2



```


# Comment
- This left the minimum count at 3
Which will break the central limit theorem but not the saddle point assumptions.


```{r}

be.m3 <- glm(cbind(Females,Males) ~  Party, family=binomial, data=belection_2)

summary(be.m3)

anova(be.m3, test = "Chisq")

exp( coef(be.m3) )



tab <- data.frame(GoF.Statistic=c(be.m3$deviance))
tab$DF <- be.m3$df.residual
tab$P.Value <- pchisq(tab$GoF, df=tab$DF, lower.tail=FALSE)
row.names(tab) <- c("Deviance"); print(tab, digits=3)



print("")
print("Residuals compared against the Proportion of females")
cbind(qres, belection_2$Females/(belection_2$Males + belection_2$Females))



```

# Comments
-  We can see from this output that the deviance is a bit higher than the degrees of freedom, indicating overdispersion

The comparison with the proportion of females with the residuals shows that the belection high residuals (2.7 and -2.91) do not come from particularly low values, one has a female proportion of 0.32 and the other 0.069 which are neither highest or lowest, so it is not from an outline. 

Only utilizing the deviance goodness of fit, as the only the saddle point approximation is necessary for this.

With a high p-value from the g-test value we find an adequate correlation for goodness of fit between the observed and expected.



Because the ratio between the deviance and the degrees of freedom is beyond 1.1 will test again with a quasi-binomial. 
The over dispersion would indicate a breaking an assumption of the binomial identity. 

# Reattempting with quasi binomial:

```{r}

be.m4 <- glm(cbind(Females,Males) ~  Party, family=quasibinomial, data=belection_2)

summary(be.m4)
anova(be.m4, test = "F") # F for quasibiomial 

c( deviance(be.m4), df.residual(be.m4 ))

#dispersion parameter
deviance(be.m4) / df.residual(be.m4)

tab <- data.frame(GoF.Statistic=c(be.m4$deviance))
tab$DF <- be.m4$df.residual
tab$P.Value <- pchisq(tab$GoF, df=tab$DF, lower.tail=FALSE)
row.names(tab) <- c("Deviance"); print(tab, digits=3)


```
The dispersion parameter is small at  1.187963, so will use just the binomial model.


#Diagnostic of the models:

be.m3 -> final binomial model

```{r}
library(statmod)
qres <- qresid(be.m3); qqnorm(qres, las=1); abline(0, 1)
scatter.smooth( qres~fitted(be.m3), las=1, main="Residuals vs fitted",
xlab="Fitted value", ylab="Quantile residual")
```

Plot of fitted values against residuals does not show a strong trend, which means that the model fits quite well. 
The qq plot confirms that the Random Component has been chosen adequately. 


```{r}
cd <- cooks.distance(be.m3)
plot(cd, type="h",ylab="Cook's distance", las=1)
obs <- which.max(cd)
cd[obs]
```

The value of Cook's distance for Observation 18 is much larger than any others, but the observation is not identified as significantly influential. To demonstrate, we fit the model without Observation 18, then compare the estimated coefficients:

```{r}
bem3.infl <- update(be.m3, subset = (-obs))
coef(bem3.infl)

coef(be.m3)

```
The changes are not substantial (only PartyLabour changes).

Check the systematic component of the model:

To examine the link function, an informal check is to plot the working responses against η^i.
If the link function is appropriate, the plot should be roughly linear.


```{r}

eta <-be.m3$linear.predictor
#eta

z <-resid(be.m3,type="working") + eta
#z

plot(z~eta, las=1,xlab="Linear predictor,eta",
     ylab="Working residuals,z")


abline(0,1,col="blue")

```

